Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-01 application.tex
regular update

We remark that it is possible, in principle, to satisfy $\beta <
\beta_\mathrm{crit}$ by choosing a sufficiently large $\khat$,
equivalently a sufficiently small perturbation lengthscales. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. We see that in order for the
fundamental VSI to operate in the inner disk (few AU), one needs to
consider large values of $\khat$.

However, our linear
calculations indicate growth rates eventually decay as one increases
$\khat$. This, together with our finding of high $\khat$ modes may be
affected by vertical boundary conditions, suggest the VSI does not
operate in the inner few AU in protoplanetary disks where thermal
relaxation is entirely due to radiation \citep[cf.][]{stoll14}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mink} 
  \caption{The minimum value of the perturbation wavenumber $\khat$ in
    the MMSN such that the associated dimensionless thermal
    relaxation time  $\beta$ is  less than the critical timescale
    $\beta_\mathrm{crit}$. 
    \label{mmsn_bcrit_bcool_mink}}   
\end{figure}

@@

We first compare $\beta$ at these radii in Fig. \ref{beta_compare}. In
the optically thin outer disk, $\beta$ is independent of $z$ and
$\beta<\beta_\mathrm{crit}$ throughout the disk column. We thus expect
the VSI to operate at $r=50AU$. In the inner disk, $\beta$ depends on 
height such that $\beta\lesssim\beta_\mathrm{crit}$ only for 
$|z|\gtrsim 2H$, which suggest that the VSI will be restricted
restricted to the upper and lower layers in the inner disk.

@@

Finally, we compare eigenfunctions for the vertical velocity at these
radii in Fig. \ref{mmsn_eigenvz}. We compare cases with similar growth
rates. At $r=50$AU we have vertical motion throughout the disk column
because the thermal timescale is less than critical. By contrast, at
$r=2$AU there is little vertical motion in $|z|\lesssim 2H$. 
Considering this result with Fig. \ref{beta_compare} suggest 
that the critical thermal timescale gives an indication of where the 
VSI operates even when $\beta$ depends on $z$.

>>>>

We remark that it is possible to satisfy $\beta <
\beta_\mathrm{crit}$ by choosing a sufficiently large $\khat$, or 
equivalently a sufficiently small perturbation lengthscale. We
illustrate this by plotting at each radius the value of $\khat$ such
that $\beta = \beta_\mathrm{crit}$. In the inner few AU of the MMSN,
one needs to consider large values of $\khat$ and/or regions away from
the disk midplane.

% However, our linear
% calculations indicate growth rates eventually decay as one increases
% $\khat$. This, together with our finding of high $\khat$ modes may be
% affected by vertical boundary conditions, suggest the VSI does not
% operate in the inner few AU in protoplanetary disks where thermal
% relaxation is entirely due to radiation \citep[cf.][]{stoll14}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_mink} 
  \caption{The minimum perturbation wavenumber $\khat$ in
    the MMSN such that the associated dimensionless thermal
    relaxation time  $\beta$ at $z=0$ (thick lines) and $z=3H$ (thin
    lines) is less than the critical timescale $\beta_\mathrm{crit}$.  
    \label{mmsn_bcrit_bcool_mink}}   
\end{figure}

@@

We first compare $\beta$ at these radii in Fig. \ref{beta_compare}. In
the optically thin outer disk, $\beta$ is independent of $z$ and
$\beta<\beta_\mathrm{crit}$ throughout the disk column. We thus expect
the VSI to operate at $r=50$AU. In the inner disk, $\beta$ depends on 
height such that $\beta\lesssim\beta_\mathrm{crit}$ only for 
$|z|\gtrsim 2H$, which suggest that the VSI will be restricted
restricted to the upper and lower layers in the inner disk.

@@

Finally, we compare eigenfunctions for the vertical velocity at these
radii in Fig. \ref{mmsn_eigenvz}. We compare cases with similar growth
rates. At $r=50$AU we have vertical motion throughout the disk column
because the thermal timescale is less than critical. By contrast, at
$r=2$AU there is little vertical motion in $|z|\lesssim 2H$. 
Considering this result with Fig. \ref{beta_compare} suggests, but does
not prove, that the critical thermal timescale gives an indication of
where the VSI operates even when $\beta$ depends on $z$.

### 2015-03-01 numerical.tex
new subsection growth time vs rad for mmsn
regular update

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in \S\ref{iso_poly}.

However, for $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. Our radially-local approximation is not valid for 
these small wavenumber perturbations and we do not consider such modes
further.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvector_iso_kx1} 
  \includegraphics[width=\linewidth]{figures/eigenvector_iso_kx60}
  \caption{Eigenfunction of the fundamental VSI for perturbation
    wavenumber $\khat=1$ (top) and $\khat=20\pi$ (bottom). The real 
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorW_beta0d1} 
  \caption{Numerically-calculated eigenfunction of the fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta = 0.01$
    (top) and $\beta=0.1$ (bottom). The
    perturbation radial wavenumber is $\khat=30$. The
    real (imaginary) part of $W$ is plotted as the solid (dotted)
    line.
    \label{relax_eigenW_num}}  
\end{figure}

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental VSI mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero for  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critical thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.

>>>>

Fig. \ref{lowfreq_eigenfunc} compares the fundamental VSI eigenfunction
$W$ for $\khat=1$ and $\khat=20\pi$. The latter value, corresponding
to $k_x= 200\pi/r$, was used by \cite{mcnally14}. We find $W$ is
similar for all $\khat\gtrsim10$ and is proportional to $z$, as
expected for the fundamental mode according to explicit solutions
developed in \S\ref{iso_poly}. For completeness we show the
corresponding eigenfunctions in vertical velocity in
Fig. \ref{lowfreq_eigenfuncvz}, which show that the fundamental mode
consists of zero nodes in $\delta v_z$.

For $\khat\lesssim 10$, e.g. $\khat=1$ in 
Fig. \ref{lowfreq_eigenfunc}, the perturbation magnitude near the
boundaries begin to dominate. For even smaller $\khat\lesssim 1$ (not
shown) we find the perturbation is almost entirely concentrated at the
boundaries. Our radially-local approximation is not valid for 
these small wavenumber perturbations and we do not consider such modes
further.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvector_iso_kx1} 
  \includegraphics[width=\linewidth]{figures/eigenvector_iso_kx60}
  \caption{Eigenfunction $W$ of the fundamental VSI for perturbation
    wavenumber $\khat=1$ (top) and $\khat=20\pi$ (bottom). The real 
    (imaginary) parts of $W$ are plotted as solid (dotted) lines. The
    eigenfunction has been normalized such that $\imag(W)=0$ at the
    vertical boundaries and $\mathrm{max}|W|=1$.
    \label{lowfreq_eigenfunc}
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_iso_kx1} 
  \includegraphics[width=\linewidth]{figures/eigenvectorvz_iso_kx60}
  \caption{Same as Fig. \ref{lowfreq_eigenfunc} but for the vertical
    velocity eigenfunctions.
    \label{lowfreq_eigenfuncvz}
  }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_beta0d01} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorvz_beta0d1} 
  \caption{Numerically-calculated fundamental VSI
    mode in a nearly vertically isothermal disk ($\Gamma=1.011$) with
    $(p,q,\epsilon)=(-1.5,-1,0.05)$, evolved with $\gamma=1.4$ and a dimensionless 
    thermal relaxation timescale $\beta = 0.01$
    (top) and $\beta=0.1$ (bottom). The
    real (imaginary) part of the vertical velocity perturbation
    $\delta v_z$ is plotted as the solid (dotted)
    line. The eigenfunction has been normalized such that
    $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$.
    \label{relax_eigenW_num}}  
\end{figure}

@@

For $\gamma>\Gamma$, i.e. stably stratified disks,
Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
relaxation rapidly stabilizes the fundamental VSI mode. This is
similar to that observed for nearly vertically isothermal disks
(Fig. \ref{bcrit_compare1}). For $\gamma=1.7,\,2.0,\,2.5$, growth
rates reach zero at  $\beta\simeq0.17,\,0.083,\,0.045$, respectively.
Interestingly, these values are equal to
$\epsilon|q|/(\gamma-\Gamma)$. This suggests that the critical thermal
relaxation timescale for vertically non-isothermal disks can also be
estimated by Eq. \ref{iso_vsi_cond} but with $\gamma-1$ replaced by
$\gamma-\Gamma$.
